### 2018-03-01 submission/manuscript.Rmd
update results with pat edits

@@

***Individual Significant OR Taxa Classify Carcinoma Poorly:***
Since specific taxa were associated with a significant OR for carcinoma we tested whether they would also be good classifiers of carcinoma. For stool, the `r length(crc_OR_stool$measure)` significant taxa did no better at classifying those with normal colons versus those with carcinomas than chance [Figure 3A]. Likewise in unmathced tissue samples the `r length(crc_unmatched_OR_tissue$measure)` significant taxa were no better than an AUC of 0.5 [Figure 3B]. These results suggest that although these taxa are significantly associated with a decreased or increased OR of carcinoma, individually they are poor classifiers of disease.

***Select Community Models can Recapitulate Whole Community Models:***
Since specific taxa increased or decreased the OR for carcinoma but performed poorly as an individual classifier of carcinoma and assessing ORs for adenoma we assessed whether the overall bacterial community was better at classifying disease. Three models were tested and included a full taxa model, full OTU model, and significant OR taxa model. If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e. adenoma stool) was analyzed. We first tested three model AUCs. Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model.

For stool, all models used had similar AUCs for both adenoma and carcinoma [Figure 4]. However, the adenoma AUCs were barely better than chance [Figure 4A]. Similarily, when analyzing the tests sets that were comprised of genera data from other studies the adenoma models performed poorly [Figure 5A]. The carcinoma models performed much better, with both the all genera and singificant OR taxa only models having a similar ability to detect individuals with carcinomas [Figure 5B]. The most common genera in the top 10 most important variables, in the full genera models used to classify carcinomas, were *Bacteroides* and *Lachnospiraceae* [Figure 6A]. Regardless of sample type, mouth-associated genera were present in models for carcinomas but not consistently across studies [Figure 6]. For the full community OTU-based models, both *Bacteroides*, *Blautia*, *Lachnospiraceae*, and *Ruminococcaceae* were present in the top 10 consistently across studies [Figure 6B]. Overall, these results suggest that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations leading each individual microbe as a poor individual classifier but much better in aggregate with others.

The tissue-based models had results that were dependent on genera or OTU level data and matched or unmatched samples [Figure S2]. Similarily to stool, the significant OR taxa performed as well as both the full taxa and OTU models [Figure S3B]. When analyzing the tests sets that were comprised of genera data from other studies, all models performed poorly with one exception being carcinoma classification in unmatched tissue [Figure S4]. Unlike stool, the test sets for the significant OR taxa did not perform as well as the all genera-based models [Figure S4C]. For the full genera models built using either matched or unmatched samples, showed no consistent taxa representation in the top 10 most important model variables across study [Figure S4A & S4B]. Conversely, the OTU-based models built using either matched or unmatched samples showed more commonalities [Figure S4C & S4D]. In models built using matched samples *Lachnospiraceae*, *Fusobacteriaceae*, *Comamonadaceae*, and *Bacteroidaceae* appeared in the top 10 percent in the majority of studies [Figure S4C]. Although models built with unmatched samples also had *Lachnospiraceae* and *Bacteroidaceae*, a major difference was the presence of *Ruminococcaceae* in the top 10 of every study [Figure S4D]. This results suggests that either the colon tissue microbiota is study or person dependent or that that kit and/or other types of contamination associated with low biomass samples may be skewing the results.

@@

>>>>

#Load in AUC data for individual genera
ind_crc_stool_auc <- read_csv("../data/process/tables/ind_genera_auc_stool.csv") %>%  mutate(auc = ifelse(taxa == "Ruminococcus" | taxa == "Clostridium_XI", invisible(1-auc), invisible(auc)))
ind_crc_combined_stool_auc <- ind_crc_stool_auc %>% group_by(taxa) %>% summarise(auc = median(auc)) %>% mutate(study = "median") %>% ungroup()

ind_unmatched_tissue_crc_auc <- read_csv("../data/process/tables/ind_genera_auc_unmatched_tissue.csv") %>% mutate(auc = ifelse(taxa == "Dorea" | taxa == "Blautia", invisible(1-auc), invisible(auc)))
ind_crc_combined_unmatched_auc <- ind_unmatched_tissue_crc_auc %>% group_by(taxa) %>% summarise(auc = median(auc)) %>% 
  mutate(study = "median") %>% ungroup()

#Load in AUC data between select and full taxa models
adn_tissue_model_aucs <- 
  as.data.frame(read_csv("../data/process/tables/adn_ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
  mutate(type = rep(c("unmatched", "matched"), 2), 
         pvalue = as.numeric(pvalue)) %>% 
  gather(key = model_type, value = AUC, full_model, select_model) %>% 
  mutate(AUC = as.numeric(AUC)) %>% 
  bind_rows(
    read_csv("../data/process/tables/adn_tissue_rf_otu_random_comparison_summary.csv") %>% 
      mutate(type = c("matched", "unmatched"), 
             model_type = "full_otu", 
             train_model = study, 
             act_mean_auc = as.numeric(act_mean_auc)) %>% 
      rename(AUC = act_mean_auc) %>% 
      select(pvalue, study, train_model, type, model_type, AUC)))

adn_stool_model_aucs <- as.data.frame(read_csv("../data/process/tables/adn_ALL_genus_stool_RF_fullvsselect_pvalue_summary.csv") %>% gather(key = model_type, value = AUC, full_model, select_model) %>% 
  mutate(AUC = as.numeric(AUC), 
         pvalue = as.numeric(pvalue)) %>% 
  bind_rows(read_csv("../data/process/tables/adn_stool_rf_otu_random_comparison_summary.csv") %>% 
              mutate(model_type = "full_otu", 
                     train_model = study, 
                     act_mean_auc = as.numeric(act_mean_auc)) %>% 
              rename(AUC = act_mean_auc) %>% 
              select(pvalue, study, train_model, model_type, AUC)))

crc_tissue_model_aucs <- as.data.frame(read_csv("../data/process/tables/ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% mutate(type = "unmatched") %>% 
  bind_rows(read_csv("../data/process/tables/ALL_genus_matched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
  mutate(type = "matched")) %>% 
  gather(key = model_type, value = AUC, full_model, select_model) %>% 
  mutate(AUC = as.numeric(AUC)) %>% 
  bind_rows(
    read_csv("../data/process/tables/matched_tissue_rf_otu_random_comparison_summary.csv") %>% 
      mutate(type = "matched", 
             model_type = "full_otu", 
             train_model = study, 
             act_mean_auc = as.numeric(act_mean_auc)) %>% 
      rename(AUC = act_mean_auc) %>% 
      select(pvalue, study, train_model, type, model_type, AUC) %>% 
      bind_rows(
        read_csv("../data/process/tables/unmatched_tissue_rf_otu_random_comparison_summary.csv") %>% 
          mutate(type = "unmatched", 
                 model_type = "full_otu", 
                 train_model = study, 
                 act_mean_auc = as.numeric(act_mean_auc)) %>% 
          rename(AUC = act_mean_auc) %>% 
          select(pvalue, study, train_model, type, model_type, AUC))))

crc_stool_model_aucs <- as.data.frame(read_csv("../data/process/tables/ALL_genus_stool_RF_fullvsselect_pvalue_summary.csv") %>% gather(key = model_type, value = AUC, full_model, select_model) %>% 
  mutate(AUC = as.numeric(AUC), 
         pvalue = as.numeric(pvalue)) %>% 
  bind_rows(read_csv("../data/process/tables/stool_rf_otu_random_comparison_summary.csv") %>% 
              mutate(model_type = "full_otu", 
                     train_model = study, 
                     act_mean_auc = as.numeric(act_mean_auc)) %>% 
              rename(AUC = act_mean_auc) %>% 
              select(pvalue, study, train_model, model_type, AUC)))

# Pvalue comparisons tables of the different individual taxa and full taxa models
stool_ind_vs_full_taxa <- read_csv("../data/process/tables/stool_ind_vs_full_taxa_pvalue_summary.csv")

unmatched_tissue_ind_vs_full_taxa <- read_csv("../data/process/tables/unmatched_tissue_ind_vs_full_taxa_pvalue_summary.csv")

@@

***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma:***
 We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas. Whereas the OR was defined based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study, we generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC). This allowed us to use a more fluid relative abundance threshold for defining disease status. Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e. AUC=0.50) [Figure 3A]. The taxa that performed the best included Clostridium XI, Ruminococcus, and Escherichia and even these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`. Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B]. The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`. These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.

***Combined taxa model classifies subjects better than using individual taxa:*** 
Instead of attempting to classify subjects based on individual taxa, next we generated Random Forest models that combined the individual taxa and evaluated the ability to classify as subject’s disease status. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`). For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).  For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies. Clearly, pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.

***Performance of models based on taxa relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR. Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were marginally better than 0.5 for any study[Figure 4A & S3A]. In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S3B-C]. When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed. Using the data from stool samples we found that although the AUC for 6 of 7 studies increased (mean decrease = `r round(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"])*100), digits = 2)`%), the more expansive models performed worse for 1 of the studies (decrease = `r round(abs(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"])*100)), digits = 2)`%). The overall improvement in performance was statistically significant (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Of the `r length(crc_OR_stool$measure)` taxa with significant ORs, all `r length(crc_OR_stool$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study . Similarly, using the data from unmatched tissue samples we found that the AUC for 4 out of 4 studies decreased between full versus select OR models (mean decrease = `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study. These results were surprising because it demonstrated that the ability to classify subjects could be done based on a limited characterization of the communities.

***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
The previous models were based on relative abundance data where sequences were assigned to coarse taxonomic assignments (i.e. typically genus or family level). To determine whether model performance improved with a more fine scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%. We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]). The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study. Again, this result was surprising as it indicated that a finer scale classification of the sequence data and thus a larger number of features to select from, did not yield improved classification of the subjects.

***Generalizability of taxon-based models trained on one dataset to the other datasets:*** 
Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort. For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study. We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 5]. As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort. These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.

@@

Select subset community you can do as well as OTUs or genera when using all of the information and that these models perform well when trained on one cohort and tested on another.

Why didn’t the whole community work better? Communities are patchy and higher level taxonomic information pools some of that patchiness
There isn’t one single bug associated with disease – why ? possibly because multiple bugs can carry out the same function (e.g. inflammation). Tell some examples (Esch Fuso, pepto)
Also, interesting that the loss of bacteria can be associated with disease. These appear to be commonly thought as beneficial bacteria that ferment to produce SCFAs

Adenomas suck. 
Which is kind of at odds with Baxter, but Baxter looked at lesions and also included FIT
Also, variation in how adenomas are defined – small polyps vs larger neopplasias (e.g. baxter), but we couldn’t do this for the current study because we didn’t have that level of detail and most studies are small
This isn’t such a novel result (Nich Chia)
Stool may not be the best place to look for early stage adenomas

Tissue vs. Stool
Not much overlap between stool and tissue biomarkers
Stool for diagnostic – does as well as tissue for classifying carcinomas
Carcinoma result surprising (e.g. lack of Fuso) – largely because small studies?

Missing microbes
ETBfrag, Strepto, etc.
B – is a lot of things that we might lose – good and bad; mucosal, right sided tumors? We might not see with stool

Meta analysis are important (eg. obesity, crc), we need more. We learned that these are hard because…
Inaccessible datasets
Missing/vague metadata
Varying sequence quality
Small datasets

Looking forward
Potential for using microbiome as a diagnostic tool
We’re excited that we were able to validate a set of biomarkes across multiple studies
Found a phenotype we can relate to the microbiome, it’s replicable and it’s strong

### 2018-03-01 submission/manuscript.tex
update results with pat edits

\textbf{\emph{Individual Significant OR Taxa Classify Carcinoma
Poorly:}} Since specific taxa were associated with a significant OR for
carcinoma we tested whether they would also be good classifiers of
carcinoma. For stool, the 8 significant taxa did no better at
classifying those with normal colons versus those with carcinomas than
chance {[}Figure 3A{]}. Likewise in unmathced tissue samples the 3
significant taxa were no better than an AUC of 0.5 {[}Figure 3B{]}.
These results suggest that although these taxa are significantly
associated with a decreased or increased OR of carcinoma, individually
they are poor classifiers of disease.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific taxa increased or decreased the OR for
carcinoma but performed poorly as an individual classifier of carcinoma
and assessing ORs for adenoma we assessed whether the overall bacterial
community was better at classifying disease. Three models were tested
and included a full taxa model, full OTU model, and significant OR taxa
model. If no taxa were significant after multiple comparison correction
then no model for that specific grouping (i.e.~adenoma stool) was
analyzed. We first tested three model AUCs. Next, the all genera models
and any significant OR taxa models were tested across all studies that
were not used to train the model.

For stool, all models used had similar AUCs for both adenoma and
carcinoma {[}Figure 4{]}. However, the adenoma AUCs were barely better
than chance {[}Figure 4A{]}. Similarily, when analyzing the tests sets
that were comprised of genera data from other studies the adenoma models
performed poorly {[}Figure 5A{]}. The carcinoma models performed much
better, with both the all genera and singificant OR taxa only models
having a similar ability to detect individuals with carcinomas {[}Figure
5B{]}. The most common genera in the top 10 most important variables, in
the full genera models used to classify carcinomas, were
\emph{Bacteroides} and \emph{Lachnospiraceae} {[}Figure 6A{]}.
Regardless of sample type, mouth-associated genera were present in
models for carcinomas but not consistently across studies {[}Figure
6{]}. For the full community OTU-based models, both \emph{Bacteroides},
\emph{Blautia}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} were
present in the top 10 consistently across studies {[}Figure 6B{]}.
Overall, these results suggest that multiple microbes could act as the
inflammatory stimulus needed to exacerbate mutations leading each
individual microbe as a poor individual classifier but much better in
aggregate with others.

The tissue-based models had results that were dependent on genera or OTU
level data and matched or unmatched samples {[}Figure S2{]}. Similarily
to stool, the significant OR taxa performed as well as both the full
taxa and OTU models {[}Figure S3B{]}. When analyzing the tests sets that
were comprised of genera data from other studies, all models performed
poorly with one exception being carcinoma classification in unmatched
tissue {[}Figure S4{]}. Unlike stool, the test sets for the significant
OR taxa did not perform as well as the all genera-based models {[}Figure
S4C{]}. For the full genera models built using either matched or
unmatched samples, showed no consistent taxa representation in the top
10 most important model variables across study {[}Figure S4A \& S4B{]}.
Conversely, the OTU-based models built using either matched or unmatched
samples showed more commonalities {[}Figure S4C \& S4D{]}. In models
built using matched samples \emph{Lachnospiraceae},
\emph{Fusobacteriaceae}, \emph{Comamonadaceae}, and
\emph{Bacteroidaceae} appeared in the top 10 percent in the majority of
studies {[}Figure S4C{]}. Although models built with unmatched samples
also had \emph{Lachnospiraceae} and \emph{Bacteroidaceae}, a major
difference was the presence of \emph{Ruminococcaceae} in the top 10 of
every study {[}Figure S4D{]}. This results suggests that either the
colon tissue microbiota is study or person dependent or that that kit
and/or other types of contamination associated with low biomass samples
may be skewing the results.

@@

>>>>

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma:}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. Whereas the OR
was defined based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study, we generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included Clostridium XI, Ruminococcus, and
Escherichia and even these had median AUC values less than 0.588.
Likewise, in unmatched tissue samples the 8 taxa with significant ORs
taxa were marginally better than one would expect by chance {[}Figure
3B{]}. The relative abundance of Dorea was the best predictor of
carcinomas and its median AUC was only 0.62. These results suggest that
although these taxa are associated with a decreased or increased OR for
the presences of carcinomas, individually, they do a poor job of
classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa:}} Instead of attempting to classify subjects based on
individual taxa, next we generated Random Forest models that combined
the individual taxa and evaluated the ability to classify as subject's
disease status. For data from stool samples, the combined model had an
AUC of 0.75, which was significantly higher than any of the AUC values
for the individual taxa (P-value \textless{} 0.033). For the full taxa
models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae} were
the most common taxa in the top 10\% mean decrease in accuracy (MDA)
across studies. Similarly, using data from the unmatched tissue samples,
the combined model had an AUC of 0.77, which was significantly higher
than the AUC values for Blautia and Weissella (P-value \textless{}
0.037). For the full taxa models using unmatched tissue,
\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies. Clearly, pooling the information
from the taxa with significant ORs results in a model that outperforms
classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are not significantly better than those based on taxa
with significant ORs:}} Next, we asked whether a Random Forest
classification model built using all of the taxa found in the
communities would outperform the models generated using those taxa with
a significant OR. Similar to our inability to identify taxa associated
with a significant OR for the presence of adenomas, the median AUCs to
classify subjects as having normal colons or having adenomas using data
from stool or tissue samples were marginally better than 0.5 for any
study{[}Figure 4A \& S3A{]}. In contrast, the models for classifying
subjects as having normal colons or having carcinomas using data from
stool or tissue samples yielded AUC values meaningfully higher than 0.5
{[}Figure 4B \& S3B-C{]}. When we compared the models based on all of
the taxa in a community to models based on the taxa with significant
ORs, the results were mixed. Using the data from stool samples we found
that although the AUC for 6 of 7 studies increased (mean decrease =
9.53\%), the more expansive models performed worse for 1 of the studies
(decrease = 0.38\%). The overall improvement in performance was
statistically significant (one-tailed paired T-test; P-value = 0.005).
Of the 8 taxa with significant ORs, all 8 were among the top 10\% most
important taxa as measured by mean decrease in accuracy, in at least one
study . Similarly, using the data from unmatched tissue samples we found
that the AUC for 4 out of 4 studies decreased between full versus select
OR models (mean decrease = 19.11\%, one-tailed paired T-test; P-value =
0.03). Of the 3 taxa with significant ORs, all 3 were among the top 10\%
most important taxa as measured by mean decrease in accuracy, in at
least one study. These results were surprising because it demonstrated
that the ability to classify subjects could be done based on a limited
characterization of the communities.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs:}} The previous models were based on relative
abundance data where sequences were assigned to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with a more fine scale classification, we
assigned sequences to operational taxonomic units (OTUs) where the
similarity among sequences within an OTU was more than 97\%. We again
found that classification models built using all of the sequence data
for a community did a poor job of differentiating between subjects with
normal colons and those with adenomas (median AUC: 0.53 {[}0.37-
0.56{]}), but did a good job of differentiating between subjects with
normal colons and those with carcinomas (median AUC: 0.71 {[}0.5-
0.9{]}). The OTU-based models performed similarly to those constructed
using the taxa with significant ORs (one-tailed paired T-test; P-value =
0.966) and those using all taxa (one-tailed paired T-test; P-value =
0.146). Among the OTUs that had the highest mean decrease in accuracy
for the OTU-based models, we found that OTUs that affiliated with all of
the 8 taxa that had a significant OR were within the top 10\% for at
least one study. Again, this result was surprising as it indicated that
a finer scale classification of the sequence data and thus a larger
number of features to select from, did not yield improved classification
of the subjects.

\textbf{\emph{Generalizability of taxon-based models trained on one
dataset to the other datasets:}} Considering the good performance of the
Random Forest models using taxa with a significant OR and using all of
the taxa, we next asked how well the models would perform when given
data from a different subject cohort. For instance, if a model was
trained using data from the Ahn study, we wanted to know how well it
would perform using the data from the Baxter study. We found the models
trained using the taxa with a significant OR all had a higher median AUC
than the models trained using all of the taxa when tested on the other
datasets {[}Figure 5{]}. As might be expected, the difference between
the performance of the modelling approaches appeared to vary with the
size of the training cohort. These data suggest that given a sufficient
number of subjects with normal colons and carcinomas, Random Forest
models trained using a small number of taxa can accurately classify
individuals from a different cohort.

@@

Select subset community you can do as well as OTUs or genera when using
all of the information and that these models perform well when trained
on one cohort and tested on another.

Why didn't the whole community work better? Communities are patchy and
higher level taxonomic information pools some of that patchiness There
isn't one single bug associated with disease -- why ? possibly because
multiple bugs can carry out the same function (e.g.~inflammation). Tell
some examples (Esch Fuso, pepto) Also, interesting that the loss of
bacteria can be associated with disease. These appear to be commonly
thought as beneficial bacteria that ferment to produce SCFAs

Adenomas suck. Which is kind of at odds with Baxter, but Baxter looked
at lesions and also included FIT Also, variation in how adenomas are
defined -- small polyps vs larger neopplasias (e.g.~baxter), but we
couldn't do this for the current study because we didn't have that level
of detail and most studies are small This isn't such a novel result
(Nich Chia) Stool may not be the best place to look for early stage
adenomas

Tissue vs.~Stool Not much overlap between stool and tissue biomarkers
Stool for diagnostic -- does as well as tissue for classifying
carcinomas Carcinoma result surprising (e.g.~lack of Fuso) -- largely
because small studies?

Missing microbes ETBfrag, Strepto, etc. B -- is a lot of things that we
might lose -- good and bad; mucosal, right sided tumors? We might not
see with stool

Meta analysis are important (eg. obesity, crc), we need more. We learned
that these are hard because\ldots{} Inaccessible datasets Missing/vague
metadata Varying sequence quality Small datasets

Looking forward Potential for using microbiome as a diagnostic tool
We're excited that we were able to validate a set of biomarkes across
multiple studies Found a phenotype we can relate to the microbiome, it's
replicable and it's strong
